---
title: "Chapter 7 Notes"
prefer-html: true
format: 
  docx: 
    toc: true
    highlight-style: arrow
    fig-format: retina
    fig-dpi: 300
    fig-height: 6
    fig-width: 6
editor: visual
execute:
  echo: fenced
  error: true
  output: true
editor_options: 
  chunk_output_type: inline
---

# Chapter 7 Notes: Spatial Analysis with US Census data

-   defines *spatial analysis* as "performance of analytic tasks that explicitly incorporate the spatial properties of a dataset."

-   Positions R as a GIS/GIS Substitute

-   **sp** and **rgeos** as old standards; **sf** as new standard

-   Chapter focuses on integrating **sf**, **tidycensus**, and **tigris**

## 7.1 Spatial Overlay

-   defines *spatial overlay* as the layering of different geographic datasets with different geometries over each other to observe patterns and relations

### aligning coordinate reference systems

-   all layers need same CRS for spatial overlay

-   **tigris** and **tidycensus** datasets default to NAD 1983 geographic reference system

    -   **sf** uses **s2** for three-dimensional overlay

        -   This can slow things down compared to projected coordinate reference system

-   Author recommends following workflow:

    1.  Download datasets planned for use

    2.  identify appropriate CRS for layers (use `crsuggest::suggest_crs()`)

    3.  transform the dataset to projected coordinate system (use `sf::st_transform()`)

    4.  Compute spatial overlay

### Identifying geometries within a metropolitan area

-   Core-based statistical areas (i.e. metropolitan/micropolitan areas) often used for regional analysis

-   "Core-based statistical areas are defined as agglomerations of counties that are oriented around a central core or cores, and have a significant degree of population interaction as measured through commuting patterns."

    -   Metropolitan areas are CBSAs with pop \> 50K

#### Example: Kansas City metropolitan area

1.  use **tigris** to get 2020 census tracts for two states in Kansas City metro area

2.  get boundary of Kansas City metro area

```{r KC-metro}
library(tigris)
library(tidyverse)
library(sf)
options(tigris_use_cache = TRUE)

# Author uses NAD83(2011) Kansas Regional Coordinate System Zone 11)

ks_mo_tracts <- map_dfr(c("KS", "MO"), ~{
  tracts(.x, cb = TRUE, year = 2020)
}) %>% 
  st_transform(8528)

kc_metro <- core_based_statistical_areas(cb = TRUE, year = 2020) %>% 
  filter(str_detect(NAME, "Kansas City")) %>% 
  st_transform(8528)

ggplot() +
  geom_sf(data = ks_mo_tracts, fill = "white", color = "grey") +
  geom_sf(data = kc_metro, fill = NA, color = "red") +
  theme_void()
```

-   the red boundary shows the outline of the Kansas City Metropolitan area; the Grey boundaries represent census tracts.

-   We can extract the data within the red boundary using spatial subsetting

### Spatial Subsets and spatial predicates

-   subsetting uses the extent of one dataset to extract features from another dataset using *co-location* defined by a *spatial predicate*

    -   Spatial subsets can be expressed with base R indexing

```{r KSC-subsetting}
kc_tracts <- ks_mo_tracts[kc_metro, ]

ggplot() +
  geom_sf(data = kc_tracts, fill = "white", color = "grey") +
  geom_sf(data = kc_metro, fill = NA, color = "red") +
  theme_void()
```

-   This returns all census tracts *intersecting* the Kansas City Metro Area boundary

    -   the default spatial predicate is `st_intersects()`

-   If we want just those tracts *within* the boundary, we can use `st_within()` with `st_filter()` (or using base index syntax as above, but including `op = st_within` argument)

```{r KCM-within}
kc_tracts_within <- ks_mo_tracts %>% 
  st_filter(kc_metro, .predicate = st_within)

# Alternative:
kc_metro2 <- kc_tracts[kc_metro, op = st_within]

ggplot() +
  geom_sf(data = kc_tracts_within, fill = "white", color = "grey") +
  geom_sf(data = kc_metro, fill = NA, color = "red") +
  theme_void()
```

## 7.2 Spatial Joins

-   Spatial Joins transfer attributes between spatial layers

    -   where a table join uses a key field to join, spatial joins use spatial predicates

        -   **sf** uses `st_join()` for spatial joins

### Point-in-polygon spatial joins

-   Suppose we are a health data analysis

-   We need to determine \$ of residents 65+ w/o insurance in patients' neighborhoods

    -   we have patients IDs, lon, and lat for each 'patient'

```{r gainsville}
library(tidyverse)
library(sf)
library(tidycensus)
library(mapview)

gainesville_patients <- tibble(
  patient_id = 1:10,
  longitude = c(-82.308131, -82.311972, -82.361748, -82.374377, 
                -82.38177, -82.259461, -82.367436, -82.404031, 
                -82.43289, -82.461844),
  latitude = c(29.645933, 29.655195, 29.621759, 29.653576, 
               29.677201, 29.674923, 29.71099, 29.711587, 
               29.648227, 29.624037))
```

-   we need to convert the data to a simple features object

    -   `st_as_sf()` from **sf** takes a data frame or tibble with lon/lat and creates a point dataset

```{r FL-points}
gainesville_sf <- gainesville_patients %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326) %>% 
  st_transform(6440) # transform to NAD83(2011) / Florida North

mapview(
  gainesville_sf,
  col.regions = "red",
  legend = FALSE)
```

-   Next, we acquire the health insurance data

    -   we have spatial & demo data, then we pre-process

        -   we call `select()` to retain three columns - GEOID, ACS estimate, and MOE

        -   we call `st_transform()` to allign the coordinate systems

```{r alachua-data}
alachua_insurance <- get_acs(
  geography = "tract",
  variables = "DP03_0096P",
  state = "FL",
  county = "Alachua",
  year = 2019,
  geometry = TRUE
) %>% 
  select(GEOID, pct_insured = estimate,
         pct_insured_moe = moe) %>% 
  st_transform(6440)
```

-   we have spatial & demo data; now we can pre-process prior to the spatial join

    -   we call `select()` to retain three columns - GEOID, ACS estimate, and MOE

    -   we call `st_transform()` to allign the coordinate systems

```{r interactive-layering}
mapview(alachua_insurance,
        zcol = "pct_insured",
        layer.name = "% with health<br/>insurance"
        ) +
  mapview(
    gainesville_sf,
    col.regions = "red",
    legend = FALSE
  )
```

-   to perform the join, we now move on to `st_join()`

    -   `st_join()` defaults to the `st_intersects()` predicate

```{r spatial-join}
patients_joined <- st_join(
  gainesville_sf,
  alachua_insurance
)
```

-   The resultant dataframe gives us the patients along with the rate (and moe) of uninsured in their respective census tracts

    -   be wary of the *ecological fallacy* - do not infer the individual is represented by group characteristics.

### Spatial joins and group-wise spatial analysis

-   we may also be interested in looking at areas within other areas

    -   i.e. neighborhoods within metropolitan areas

-   This can be achieved using *polygon-on-polygon spatial* joins

#### Spatial join data setup

-   We want to look at distribution of census tracts by Hispanic population for four largest metro areas in Texas

    -   We will use `B01003_001` from the 1-year ACS to get CBSA population data, and the geometry

```{r}
library(tidycensus)
library(tidyverse)
library(sf)

# CRS: NAD83(2011) / Texas Centric Albers Equal Area
tx_cbsa <- get_acs( # get the data
  geography = "cbsa",
  variables = "B01003_001",
  year = 2019,
  survey = "acs1",
  geometry = TRUE
) %>% filter(str_detect(NAME, "TX")) %>% # get those w/ names containing texas
  slice_max(estimate, n = 4) %>% # get the highest 4 estimates
  st_transform(6579) # transform CRS
```

-   We can now pull down ACS 5-year data from 2019 for Texas tracts

```{r}
pct_hispanic <- get_acs(
  geography = "tract",
  variables = "DP05_0071P",
  state = "TX",
  year = 2019,
  geometry = TRUE
) %>% 
  st_transform(6579)
```

-   We can now keep only the tracts we are interested in using a spatial join

#### Computing and visualizing the spatial join

-   with `st_join()`,dataset `x` retains geometry and gains attributes from dataset `y` based on spatial predicates

    -   `suffix` passes a suffix to be appended to columns with same names

    -   `left = FALSE` requests inner join

```{r}
hispanic_by_metro <- st_join(
  pct_hispanic,
  tx_cbsa,
  join = st_within,
  suffix = c("_tracts", "_metro"),
  left = FALSE
)
```

-   We can now perform group-wise data visualization and analysis across metro areas

```{r metro-facet}
hispanic_by_metro %>% 
  mutate(NAME_metro = str_replace(NAME_metro, ", TX Metro Area", "")) %>% 
  ggplot() +
  geom_density(aes(x = estimate_tracts), color = "navy", fill = "navy",
               alpha = 0.4) +
  theme_minimal() +
  facet_wrap(~NAME_metro) +
  labs(title = "Distribution of Hispanic/Latino population by Census tract",
       subtitle = "Largest metropolitan area in Texas",
       y ="Kernel density estimate",
       x = "Percent Hispanic/Latino in Census tract")
```

-   We can also generate group-wise summary statistics

```{r median-group-wise}
median_by_metro <- hispanic_by_metro %>% 
  group_by(NAME_metro) %>% 
  summarize(median_hispanic = median(estimate_tracts, na.rm = TRUE))

median_by_metro
```

-   Note: the above `summarize()` operation also dissolves group geographies

## 7.3 Small area time-series analysis
