---
title: "Chapter 2 Notes"
format: 
  docx: 
    toc: true
    highlight-style: arrow
editor: visual
execute:
  echo: fenced
  error: true
  output: true
editor_options: 
  chunk_output_type: inline
---

# Chapter 2: An introduction to tidycensus

**tidycensus** package has two goals:

1.  make ***tidy*** census data
2.  streamline the spatial data wrangling by including *geometry* requests

**tidycensus** focuses on subset of Census data, via core functions:

-   `get_decennial()` returns data from 2000, 2010, and 2020 Decennial Census APIs

-   `get_acs()` returns data from American Community Survey

    -   1-year back to 2005

    -   5-year back to 2005-2009

-   `get_estimates()` access population estimates apis for yearly estimates of population characteristics

    -   geometries include:

        -   state

        -   county

        -   metropolitan area

    -   estimates:

        -   births

        -   deaths

        -   migration rates

-   `get_pums()` access Public Use Microdata Samples

-   `get_flows()` Migration flows API access, including info

## 2.1 Getting started with tidycensus

```{r tidycensus-setup}
# install and/or load tidycensus package and set API key
#install.packages("tidycensus")
library(tidycensus)
library(tidyverse)
# census_api_key("insert api key here", install = TRUE)
```

### Decennial Census

`get_decennial()` function calls

```{r get_decennial}
total_pop_10 <- get_decennial(
  geography = "state", # aggregate by state
  variables = "P001001",
  year = 2010 # year we want data for
) 
```

-   pulls from Summary File 1 by default, using default argument `sumfile = "sf1"`

-   Summary File 1 exists for 2000 and 2010 Censuses, and contains core demo info for Census Geographies

-   Summary File 2 contains info on population range and housing units

-   Summary File 3 and 4 contain the Long Form data for 2000; these data sets are in American Community Survey sets from 2005 onward

### ACS

`get_acs()` gets data from American Community Survey.

ACS collects a lot more variables not included in the short form questionnaire of the decennial Census.

```{r ACS}
born_in_mexico <- get_acs(
  geography = "state",
  variables = "B05006_150",
  year = 2020
)
born_in_mexico
```

If the year isn't specified, `get_acs()` defaults to most recent five-year ACS sample

defaults to 5-year ACS, but can be changed to 1 year using `survey = "acs1"`

```{r mexican-born-1yr}
born_in_mexico_1yr <- get_acs(
  geography = "state",
  variables = "B05006_150",
  survey = "acs1",
  year = 2019
)
born_in_mexico_1yr
```

Variables from ACS detailed tables, data profiles, summary tables, comparison profile, and supplemental estimates are available through the function.

-   auto-detects where to look based on variable name

-   alternatively, supply table name to get data for all variables in table

***Example: Table B01001 ( Sex by Age)***

```{r B01001}
age_table <- get_acs(
  geography = "state",
  table = "B01001",
  year = 2020
)
age_table
```

## 2.2 Geography and variables in tidycensus

`geography` parameter in `get_acs()` and \`get_decennial() requests data aggregated to common Census units

-   Census blocks are only available in `get_decennial()`

-   Only geographies available in 2000 are `"state"`, `"county"`, `"county subdivision"`, `"tract"`, `"block group"`, and `"place"`

-   the geographies have to be typed exactly per the table in the documentation

    -   core-based statistical areas and zip code tabulation areas have `"cbsa"` and `"zcta"` as aliases

```{r cbsa-demo}
cbsa_population <- get_acs(
  geography = "cbsa",
  variables = "B01003_001",
  year = 2020
)
cbsa_population
```

### Geographic subsets

**tidycensus** supports geographic subsetting if supported by the API

***Example: Wisconsin income analysis***

```{r wi-income}
wi_income <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "WI",
  year = 2020
)
wi_income
```

Smaller geographies (census tracts, block groups, etc) can be subset by county.

***Example: Dane County, WI***

```{r dane-income}
dane_income <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "WI",
  county = "Dane",
  year = 2020
)
dane_income
```

**NOTE:** 5-year covers down to block group; 1-year only down to geographies with ≥ 65,000 pop

## 2.3 Searching for variables in tidycensus

Variable IDs are a PITA

**tidycensus** has `load_variables()` to help with searching

-   two required arguments:

    1.  `year`: the year or end-year for census dataset or ACS sample

    2.  `dataset`: the dataset name

        1.  `"sf1"` = Summary File 1 for 2000 or 2010

        2.  `"sf2"` = Summary File 2 for 2000 or 2010

        3.  `"sf3"` = Summary File 3 for 2000

        4.  `"sf4"` = Summary File 4 for 2000

        5.  `"pl"` = PL-94171 Redistricting dataset for 2020

        6.  `"acs1"` 1-year ACS

        7.  `"acs5"` 5-year ACS

            -   ACS Data Profile, Summary Tables, and Comparison Profile requests require the suffixes `/profile`, `/summary`, `/cprofile` respectively

        8.  `"acsse"` = ACS Supplemental Estimates variables

    3.  Users can specify argument `cache = TRUE` to store data in user cache

```{r examples}
v16 <- load_variables(2016, "acs5", cache = TRUE)
```

## 2.4 Data structure in tidycensus

**tidycensus** is built to function in the tidyverse; Wickham's definition of "tidy" data is as follows:

1.  Each observation forms a row
2.  Each variable forms a column
3.  Each observational unit forms a table

**tidycensus** returns tibbles of decennial Census data tidied as follows:

-   `GEOID` - the Census ID code giving unique ID to the geographic unit

-   `NAME` representing a descriptive name of the unit

-   `variable` contain info on Census variable name for the row

-   `value` the data value for each unit-variable combination.

    -   in ACS data, `value` is replaced with `estimate` and `moe` which represents the margin of error around the estimate

Census or ACS "variables" are instead *characteristics* of enumeration units in **tidycensus**

-   default setting is `output = "tidy"`

-   rows represent data for unique spatial unit-variable combination

***Example: hhinc demo***

```{r hhcin-demo}
hhinc <- get_acs(
  geography = "state",
  table = "B19001",
  survey = "acs1",
  year = 2016
)
hhinc
```

Row 1 demonstrates the estimate for the state-characteristic pair Alabama-B19001_001, with its associated moe

For a 'wide' table, use `output = "wide"`

```{r hhinc_wide}
hhinc_wide <- get_acs(
  geography = "state",
  table = "B19001",
  survey = "acs1",
  year = 2016,
  output = "wide"
)
hhinc_wide
```

### Understanding GEOIDs

-   For core Census hierarchy (Census block -\> block group -\> census tract -\> county -\> state -\> division -\> region -\> nation) GEOID identifies specific unit AND unit's parent geographies

***Example: Cimarron County, Oklahoma GEOIDs***

```{r}
cimarron_blocks <- get_decennial(
  geography = "block",
  variables = "H1_001N",
  state = "OK",
  county = "Cimarron",
  year = 2020,
  sumfile = "pl"
)
cimarron_blocks
```

For the GEOID, 40.025.950100.1.501 (periods added for clarity), it can be read as follows

-   first two digits (**40)** are the State FIPS code for Oklahoma

-   digits three through 5 (**025)** correspond to Cimarron County within Oklahoma

    -   To look at counties across the whole US, the 5 digits together uniquely identify a county - so Oklahoma's Cimarron County can be read as **40025**

-   next six digits represent the Census tract (**950100)** - in this case, Census Tract 9501, with zeroes padding the right.

-   Twelfth digit **(1)** represents the parent block group of the Census block

    -   A census tract will have up to, but no more than, 9 census block groups

-   The final three digits (**501)** represent the individual census block.

    -   The block's name is a combination of this ID and the parent block group digit (**1501)**

For geographies outside the core hierarchy, will only contain IDs of parent geographies they fully nest within.

cbsa's and zcta's will have their own unique GEOIDs independent of other aggregation levels.

### Renaming variable IDs

passing a named vector to `variables` parameter will feturn desired names rather than Census variable IDs

```{r georgia-demo}
ga <- get_acs(
  geography = "county",
  state = "Georgia",
  variables = c(medinc = "B19013_001",
                medage = "B01002_001"),
  year = 2020
)
ga

ga_wide <- get_acs(
  geography = "county",
  state = "Georgia",
  variables = c(medinc = "B19013_001",
                medage = "B01002_001"),
  year = 2020,
  output = "wide"
)
ga_wide
```

## 2.5 Other Census Bureau datasets in tidycensus

-   **tidycensus** does not grant access to all datasets from Census API

    -   **censusapi** provides more general api access

    Two other datasets with custom functions are `get_estimates()` and `get_flows()` to access Population Estimates and ACS Migration Flows, respectively.

### Using `get_estimates()`

**Population Estimates Program (PEP)** provides yearly estimates based on projections drawn from birth, death, and migration rates from the most recent Census.

It does provide access to indicators used for projection

-   indicators can be specified using the `product` argument, either individually or in bulk.

    -   population change estimates are pulled with `"components"`

***Example: Queens County population estimate components***

```{r pep_components}
queens_components <- get_estimates(
  geography = "county",
  product = "components",
  state = "NY",
  county = "Queens",
  year = 2019
  )
queens_components
```

The variables returned above include:

-   BIRTHS: raw count of births from July 1 2018 to July 1 2019

-   DEATHS: raw count of births from July 1 2018 to July 1 2019

-   DOMESTICMIG: Net domestic migration in period

-   INTERNATIONALMIG: Net international migration in period

-   NATURALINC: Natural increase in period (Birth - Death)

-   NETMIG: Net migration in period

-   RBIRTH: Crude birth rate in period (per 1000)

-   RDEATH: Crude death rate in period (per 1000)

-   RDOMESTICMIG: Crude Domestic Migration Rate

-   RINTERNATIONALMIG: Crude International Migration Rate

-   RNATURALINC: Crude Natural Increase Rate

-   RNETMIG: Crude Net Migration Rate

The `"characteristics"` product lets users get unique demographic info, esp. when using `breakdown` argument with the arguments `"AGEGROUP"`, `"RACE"`, `"SEX"`, or `"HISP"`

The `breakdown_labels = TRUE` returns informative labels

```{r}
louisiana_sex_hisp <- get_estimates(
  geography = "state",
  product = "characteristics",
  breakdown = c("SEX", "HISP"),
  breakdown_labels = TRUE,
  state = "LA",
  year = 2019
)
louisiana_sex_hisp
```

### Using get_flows()

`get_flows()` pulls down info on in- and out-migration for states, counties, and metropolitan areas.

-   pulls for a given geography, using data from a given 5-year ACS sample.

***Example: Honolulu County, Hawaii***

```{r example-flow}
honolulu_migration <- get_flows(
  geography = "county",
  state = "HI",
  county = "Honolulu",
  year = 2019
)
honolulu_migration
```

`get_flows()` also includes migration flow mapping functionality, discussed later

## 2.6 Debugging tidycensus errors

**`tidycensus`** carries through census API errors where possible, and tries to translate common errors.

***Example: mis-typed variable error***

```{r}
state_pop <- get_decennial(
  geography = "state",
  variables = "P01001",
  year = 2010
)
```

Note in console, error: `Error : Your API call has errors.  The API message returned is error: error: unknown variable 'P01001'. Using Census Summary File 1`

For an example where a passed geography is not present in the given dataset

***Example: Unavailable Geography***

```{r}
cbsa_ohio <- get_acs(
  geography = "cbsa",
  variables = "DP02_0068P",
  state = "OH",
  year = 2019
)
```

**`tidycensus`** includes the `show_call` parameter that prints out the API call made

```{r}
cbsa_bachelors <- get_acs(
  geography = "cbsa",
  variables = "DP02_0068P",
  year = 2019,
  show_call = TRUE
)
```

## Exercises

1.  Review the available geographies in tidycensus from the geography table in this chapter. Acquire data on median age (variable `B01002_001`) for a geography we have not used.

    ```{r}
    california <- get_acs(
      geography = "zcta",
      variables = "B01002_001",
      show_call = FALSE)

    california
    ```

2.  Use the `load_variables()` function to find a variable that interests you that we haven't used yet. Use `get_acs()` to fetch data from the 2016-2020 ACS for counties in the state where you live, where you have visited, or where you would like to visit.

    ```{r}
    #summary_files(2020)
    load_variables(2020, "dp")

    los_angeles_u_5 <- get_acs(
      geography = "tract",
      state = "CA",
      county = "Los Angeles",
      variables = c(male_u_5 = "B01001_003", 
                    female_u_5 = "B01001_027",
                    total_pop_est = "B01001_001"),
      summary_var = "B01001_001"
    )

    los_angeles_u_5 <- get_decennial(
      geography = "tract",
      state = "CA",
      county = "Los Angeles",
      variables = c(total_u_5 = "DP1_0002C"),
      sumfile = "dp"
    )
    ```

    ACS5 data for age under 5

    -   B01001_003 = Male Estimated Total Age Under 5

    -   B01001_027 = Female Estimated Total Age under 5

    -   B01001_001 = Estimated Total, regardless of age or gender

    Decennial census 2020 data for age under 5

    -   DP1_0002C = Count of total population under age of 5
